---
title: "exploring life expectancy data for MPA"
output: html_document
---

<style type="text/css">

body{ /* Normal  */
      font-size: 16px;
  }

h1.title {
  font-size: 16px;
  color: DarkRed;
}

h1 { /* Header 1 */
  font-size: 34px;
  color: DarkRed;
}

h2 { /* Header 2 */
    font-size: 18px;
  color: DarkRed;
}

</style>


```{r setup, include=FALSE, messages=FALSE}
knitr::opts_chunk$set(echo = FALSE, messages=FALSE, warnings=FALSE)
library(dplyr)
library(summarytools)
library(readxl)
library(ggplot2)
library(janitor)
library(knitr)
library(stringr)

myPlace <- getwd()

```


```{r}

raceLink   <- as.data.frame(read_excel("/mnt/projects/CCB/CCB Project/0.CCB/myCBD/myInfo/raceLink.xlsx")) 
                                                                                                   

lifeExpMPA    <- readRDS(paste0("/mnt/projects/CCB/CCB Project/0.CCB/myUpstream/lifeTables/dataOut/LTciCounty.rds")) %>%
                  left_join(raceLink,by=c("race7" = "Ethan")) %>%
                  filter(x %in% c(0,50)) %>%
                  filter(year > 2012, nyrs==5) %>%
                  mutate(GEOID = str_sub(GEOID,1,5)) %>%
                  select(year, GEOID, sex, raceName, lAge = x, population = Nx, deaths = Dx, life_expectancy = ex, LE_lci = exlow, LE_uci = exhigh)
 
 
lifeExpMPA.State <- readRDS(paste0("/mnt/projects/CCB/CCB Project/0.CCB/myUpstream/lifeTables/dataOut/LTciState.rds")) %>%
                  left_join(raceLink,by=c("race7" = "Ethan")) %>%
                  filter(x %in% c(0,50)) %>%
                  filter(year > 2012) %>%
                  mutate(GEOID = str_sub(GEOID,1,5)) %>%
                  select(year, GEOID, sex, raceName, lAge = x, population = Nx, deaths = Dx, life_expectancy = ex, LE_lci = exlow, LE_uci = exhigh)


lifeExpMPA <- bind_rows(lifeExpMPA,lifeExpMPA.State)


write.csv(lifeExpMPA, "/mnt/projects/CCB/CCB Project/0.CCB/myUpstream/lifeTables/dataOut/lifeExpMPA.csv")

#write.csv(lifeExpMPA, "G:/CCB/CCB Project/0.CCB/myUpstream/lifeTables/dataOut/lifeExpMPA.csv")


```


* discuss all cause mortality
* discuss premature mortality

* sharing this chart just to show the data we have; data table of same data readily available too of course
* this chart is for California overall; same data is available for most counties
* these data are based on five-year aggregations of data, 2014-2018; we may be able to update to 2015-2019 soon

* the charts on the left with the "0" label are life expectancy at birth
* the charts on the right are life expectancy for persons who make it to age 50
* I suggest we focus on just life expectancy at birth


```{r}

lifeExp <- mutate(lifeExp,ex=ifelse(x==50,ex+50,ex))


ggplot(data=lifeExp, aes(x=race7,y=ex)) + geom_bar(stat="identity") +
  facet_grid(rows=vars(sex),cols=vars(x)) +  geom_text(aes(label = round(ex,1)),nudge_y=-10,col="yellow") 

```


<br><br>

```{r}
tDat <- lifeExp %>% filter(x==0)  %>%
         group_by(sex) %>%
         mutate(maxLife     = max(ex),
                yearsLess    = maxLife - ex,
                percentLess = 100*yearsLess/maxLife) %>%
         select(sex,race7,ex,yearsLess,percentLess) %>%
         adorn_rounding(digits = 1)

kable(tDat)

```

