---
title: "Suicide-Homicide Trends by County"
author: "Fusion Center"
date: "July 2019"
output:
  pdf_document: default
  html_document: default
  word_document: default
---


```{r setup, include=FALSE}
knitr::opts_chunk$set( echo = FALSE, warning= FALSE, message=FALSE)  #out.width = "100%",

 library(dplyr)
 library(ggplot2)
 library(plotly)
 library(tidyr)
 library(readxl)
 library(readr) 
 library(fs)
 library(markdown)
 library(kableExtra)
 library(DT)

 myDrive <- getwd()  
 myDrive <- "F:/0.CBD"
 myPlace <- paste0(myDrive,"/myCBD") 
 upPlace <- paste0(myDrive,"/myUpstream") 

 # read file to link Cause codes to names
 gbdMap0                 <- as.data.frame(read_excel( path(myPlace,"myInfo//gbd.ICD.Map.xlsx/"), sheet="main"))   
 fullCauseList           <- gbdMap0[!is.na(gbdMap0$causeList),c("LABEL","nameOnly")] %>% arrange(LABEL)
 names(fullCauseList)[2] <- "Cause_Name"
 
 Causes                  <- c("E02","E03")  # suicides and homicides
 
 # read 3-year county data set, get cause names, select "Total" sex and just homicides and suicides
 t.dat       <- readRDS(path(myPlace,"/myData/","noSupression","datCounty_3year.RDS")) %>%
                  left_join(fullCauseList, by = c("CAUSE" = "LABEL"))          %>% 
                  filter(sex=="Total", CAUSE %in% Causes)
 
 # read (5-year county data set, get cause names, select "Total" sex and just homicides and suicides
 t.dat.Comm  <- readRDS(path(myPlace,"/myData/","noSupression","datComm.RDS"))   %>%
                  left_join(fullCauseList, by = c("CAUSE" = "LABEL"))    %>%
                  filter(sex=="Total", CAUSE %in% Causes)

 # read SDOH data set
 load(path(myPlace,"/myData/","sdohComm.R"))
 
 
```


```{r}

 cause.labs              <- c("Suicide", "Homicide")
 names(cause.labs)       <- Causes
 

 tPlot <-   ggplot(filter(t.dat,CAUSE=="E02"),aes(yearG3,aRate,group=county)) +
             geom_line(aes(color=county)) +
                         theme(axis.text.x = element_text(angle = 90, hjust = 1),strip.text.y = element_text(angle = 0)) +
   labs(title="Suicide Rate by California County, Trend 2000-02 to 2015-17") + ylab("age-adjusted Rate per 100,000")

 
 
 # tPlot <-   ggplot(t.dat,aes(yearG3,aRate,group=county)) +
 #             geom_line(aes(color=county)) +
 #             facet_grid(CAUSE~, labeller = labeller(CAUSE = cause.labs)) +
 #             theme(axis.text.x = element_text(angle = 90, hjust = 1),strip.text.y = element_text(angle = 0))

 
 
 ggplotly(tPlot)
 
 
  tPlot <-   ggplot(filter(t.dat,CAUSE=="E03"),aes(yearG3,aRate,group=county)) +
             geom_line(aes(color=county)) +
                         theme(axis.text.x = element_text(angle = 90, hjust = 1),strip.text.y = element_text(angle = 0)) +
   labs(title="Homicide Rate by California County, Trend 2000-02 to 2015-17") + ylab("age-adjusted Rate per 100,000")
  
   ggplotly(tPlot)
 

```



```{r}
# FUNCTION to make homoicide or suicide table


myTable  <- function(myCAUSE) {
  
  
tState <- filter(t.dat,county=="CALIFORNIA",yearG3=="2015-2017",CAUSE==myCAUSE)  %>%
           select(aRate,aSE)
  
  
t0  <-  t.dat %>% select(CAUSE,yearG3,county,Ndeaths,aRate,aLCI,aUCI,aSE)  %>% 
                  filter(CAUSE== myCAUSE)

tNs  <-  t0 %>% select(county,yearG3,Ndeaths) %>% spread(yearG3,Ndeaths) %>%  
               rename_at(vars(-county), list(~paste0("Ndeaths ", .)))

tRates  <-  t0 %>% select(county,yearG3,aRate) %>% spread(yearG3,aRate) %>%
               mutate(`Percent Change` = 100*(`2015-2017` - `2006-2008`) / `2006-2008` ) %>%
               rename_at(vars(-county,-`Percent Change`), list(~paste0("Rate ", .)))

tStats  <- t0 %>%  select(county,yearG3,aSE) %>% spread(yearG3,aSE) %>%
                   rename_at(vars(-county), list(~paste0("SE ", .)))
  
tStats <- full_join(tStats,tRates,by="county") %>% 
          mutate(zdif = (`Rate 2015-2017` - `Rate 2006-2008`  ) / sqrt(`SE 2015-2017`^2 + `SE 2006-2008`^2),
                pValue = 1-pnorm(abs(zdif)) ,
                zState = (`Rate 2015-2017` -  tState$aRate) / (sqrt(`SE 2015-2017`^2 + tState$aSE^2)),
                pState = 1-pnorm(abs(zState)),
                `County to State 2015-2017`=ifelse(pState < 0.05,ifelse(`Rate 2015-2017` > tState$aRate,"Higher","Lower"),"No Diff")
                ) 

SEcolumns <- names(tStats)[grep("SE",names(tStats))]
tStats <- tStats %>% select(-SEcolumns,-zdif,-zState,-pState)

UCI  <-  t0 %>% select(county,yearG3,aUCI) %>% spread(yearG3,aUCI) %>%
                select(county,up06_08=`2006-2008`,up15_17=`2015-2017`)
LCI  <-  t0 %>% select(county,yearG3,aLCI) %>% spread(yearG3,aLCI) %>%
                select(county,low06_08=`2006-2008`,low15_17=`2015-2017`)
CIs  <-  full_join(UCI,LCI,by="county") %>% select(county,low06_08,up06_08,low15_17,up15_17)

tWork <<- full_join(tNs,tStats,by="county") %>% full_join(CIs,by="county")
temp <- names(tWork)

brks <- c(0.01,.05,1)
clrs <- c("red","pink","white","green")


 datatable(tWork, rownames = FALSE,options = list(pageLength=15) ) %>%
     formatRound(columns=temp[c(2:7)], digits=0) %>%
     formatRound(columns=temp[c(8:13,17:20)], digits=1) %>%
     formatRound(columns=temp[c(14)], digits=2) %>%
     formatRound(columns=temp[c(15)], digits=3) %>% 
     formatStyle(columns=temp[c(4,7)],backgroundColor = 'lightblue') %>%
     formatStyle(columns=temp[c(10,13)],backgroundColor = 'gray') %>%
     formatStyle(columns=temp[c(15)],backgroundColor = styleInterval(brks, clrs))  
  
 }

```

<br><br><br>

## HOMICIDES - County Level
```{r}
myTable("E03")
```


* summary of county homicide rates compared to state rate
```{r}
tM <- table(tWork$`County to State 2015-2017`)
tMsum <- sum(tM)
tM <- c(tM,total=tMsum)

bind_rows(tM,100*tM/tMsum)

## WRITE
write_csv(tWork,"forJN_1_Homicide_County.csv")
```

<br><br><br>

## SUICIDES - County Level
```{r}
myTable("E02")
```


* summary of county suicide rates compared to state rate
```{r}
tM <- table(tWork$`County to State 2015-2017`)
tMsum <- sum(tM)
tM <- c(tM,total=tMsum)
bind_rows(tM,100*tM/tMsum)

## WRITE
write_csv(tWork,"forJN_2_Suicide_County.csv")

```


<br><br><br>

## HOMICIDES - Community Level


```{r}


## state rate and SE
tState <- filter(t.dat,county=="CALIFORNIA",yearG3=="2015-2017",CAUSE=="E03")  %>%
           select(aRate,aSE)



t0  <- t.dat.Comm %>% select(CAUSE,yearG5,county,comName,Ndeaths,aRate,aLCI,aUCI,aSE) %>% 
                      filter(CAUSE== "E03")

tNs  <-  t0 %>% select(county,comName,yearG5,Ndeaths) %>% spread(yearG5,Ndeaths) %>%  
               rename_at(vars(-county,-comName), list(~paste0("Ndeaths ", .)))

tRates  <-  t0 %>% select(county,comName,yearG5,aRate) %>% spread(yearG5,aRate) %>%
               mutate(`Percent Change` = 100*(`2013-2017` - `2008-2012`) / `2008-2012` ) %>%
               rename_at(vars(-county,-comName, -`Percent Change`), list(~paste0("Rate ", .)))

tStats  <- t0 %>%  select(county,comName,yearG5,aSE) %>% spread(yearG5,aSE) %>%
                   rename_at(vars(-county,-comName), list(~paste0("SE ", .)))
  
tStats <- full_join(tStats,tRates,by=c("county","comName")) %>% 
          mutate(zdif = (`Rate 2013-2017` - `Rate 2008-2012`  ) / sqrt(`SE 2013-2017`^2 + `SE 2008-2012`^2),
                pValue = 1-pnorm(abs(zdif)),
                zState = (`Rate 2013-2017` -  tState$aRate) / (sqrt(`SE 2013-2017`^2 + tState$aSE^2)),
                pState = 1-pnorm(abs(zState)),
                `Community 13-17 Compared to State 15-17`=ifelse(pState < 0.05,ifelse(`Rate 2013-2017` > tState$aRate,"Higher","Lower"),"No Diff")
                ) 

SEcolumns <- names(tStats)[grep("SE",names(tStats))]
tStats <- tStats %>% select(-SEcolumns,-zdif,-zState,-pState)

tWork <- full_join(tNs,tStats,by=c("county","comName")) 
temp <- names(tWork)

brks <- c(0.01,.05,1)
clrs <- c("red","pink","white","green")


datatable(tWork, rownames = FALSE,options = list(pageLength=15) ) %>%
    formatRound(columns=temp[c(5:5)], digits=1) %>%
    formatRound(columns=temp[c(7)], digits=2) %>%
    formatRound(columns=temp[c(8)], digits=3) %>% 
    formatStyle(columns=temp[c(8)],backgroundColor = styleInterval(brks, clrs))  


tM <- table(tWork$`Community 13-17 Compared to State 15-17`)
tMsum <- sum(tM)
tM <- c(tM,total=tMsum)
bind_rows(tM,100*tM/tMsum)


## WRITE
write_csv(tWork,"forJN_3_Homicide_Community.csv")


```

<br><br><br>



## SUICIDES - Community Level



```{r}


## state rate and SE
tState <- filter(t.dat,county=="CALIFORNIA",yearG3=="2015-2017",CAUSE=="E02")  %>%
           select(aRate,aSE)



t0  <- t.dat.Comm %>% select(CAUSE,yearG5,county,comName,Ndeaths,aRate,aLCI,aUCI,aSE) %>% 
                      filter(CAUSE== "E02")

tNs  <-  t0 %>% select(county,comName,yearG5,Ndeaths) %>% spread(yearG5,Ndeaths) %>%  
               rename_at(vars(-county,-comName), list(~paste0("Ndeaths ", .)))

tRates  <-  t0 %>% select(county,comName,yearG5,aRate) %>% spread(yearG5,aRate) %>%
               mutate(`Percent Change` = 100*(`2013-2017` - `2008-2012`) / `2008-2012` ) %>%
               rename_at(vars(-county,-comName, -`Percent Change`), list(~paste0("Rate ", .)))

tStats  <- t0 %>%  select(county,comName,yearG5,aSE) %>% spread(yearG5,aSE) %>%
                   rename_at(vars(-county,-comName), list(~paste0("SE ", .)))
  
tStats <- full_join(tStats,tRates,by=c("county","comName")) %>% 
          mutate(zdif = (`Rate 2013-2017` - `Rate 2008-2012`  ) / sqrt(`SE 2013-2017`^2 + `SE 2008-2012`^2),
                pValue = 1-pnorm(abs(zdif)),
                zState = (`Rate 2013-2017` -  tState$aRate) / (sqrt(`SE 2013-2017`^2 + tState$aSE^2)),
                pState = 1-pnorm(abs(zState)),
                `Community 13-17 Compared to State 15-17`=ifelse(pState < 0.05,ifelse(`Rate 2013-2017` > tState$aRate,"Higher","Lower"),"No Diff")
                ) 

SEcolumns <- names(tStats)[grep("SE",names(tStats))]
tStats <- tStats %>% select(-SEcolumns,-zdif,-zState,-pState)

tWork <- full_join(tNs,tStats,by=c("county","comName")) 
temp <- names(tWork)

brks <- c(0.01,.05,1)
clrs <- c("red","pink","white","green")


datatable(tWork, rownames = FALSE,options = list(pageLength=15) ) %>%
    formatRound(columns=temp[c(5:5)], digits=1) %>%
    formatRound(columns=temp[c(7)], digits=2) %>%
    formatRound(columns=temp[c(8)], digits=3) %>% 
    formatStyle(columns=temp[c(8)],backgroundColor = styleInterval(brks, clrs))  


tM <- table(tWork$`Community 13-17 Compared to State 15-17`)
tMsum <- sum(tM)
tM <- c(tM,total=tMsum)
bind_rows(tM,100*tM/tMsum)



## WRITE
write_csv(tWork,"forJN_4_Suicide_Community.csv")


```

<br><br><br>


## HOMICIDES -- INCREASES ONLY
```{r}
tab0 <- t.dat.Comm %>% select(CAUSE,yearG5,county,comName,Ndeaths,aRate,aLCI,aUCI,aSE)
myCAUSE <- "E03"



t0  <- tab0 %>% filter(CAUSE== myCAUSE)

tNs  <-  t0 %>% select(county,comName,yearG5,Ndeaths) %>% spread(yearG5,Ndeaths) %>%  
               rename_at(vars(-county,-comName), list(~paste0("Ndeaths ", .)))

tRates  <-  t0 %>% select(county,comName,yearG5,aRate) %>% spread(yearG5,aRate) %>%
               mutate(`Percent Change` = 100*(`2013-2017` - `2008-2012`) / `2008-2012` ) %>%
               rename_at(vars(-county,-comName, -`Percent Change`), list(~paste0("Rate ", .)))

tStats  <- t0 %>%  select(county,comName,yearG5,aSE) %>% spread(yearG5,aSE) %>%
                   rename_at(vars(-county,-comName), list(~paste0("SE ", .)))
  
tStats <- full_join(tStats,tRates,by=c("county","comName")) %>% 
          mutate(zdif = (`Rate 2013-2017` - `Rate 2008-2012`  ) / sqrt(`SE 2013-2017`^2 + `SE 2008-2012`^2),
                pValue = 1-pnorm(abs(zdif)) 
                ) 

SEcolumns <- names(tStats)[grep("SE",names(tStats))]
tStats <- tStats %>% select(-SEcolumns,-zdif)

tWork <- full_join(tNs,tStats,by=c("county","comName")) 
temp <- names(tWork)

brks <- c(0.01,.05,1)
clrs <- c("red","pink","white","green")


tWorkIncrease <- filter(tWork, `Percent Change` > 0)


datatable(tWorkIncrease, rownames = FALSE,options = list(pageLength=15) ) %>%
    formatRound(columns=temp[c(5:5)], digits=1) %>%
    formatRound(columns=temp[c(7)], digits=2) %>%
    formatRound(columns=temp[c(8)], digits=3) %>% 
    formatStyle(columns=temp[c(8)],backgroundColor = styleInterval(brks, clrs))  

```

<br><br><br>


# SDOH Work
### quick "exploration" of distributions of the three (plus) varibles we are "correlating"
### histograms to look at the distribution of each

 * community poverty and education from 2017 5-year ACS data, so covers 2013-2017; education is for population 25 and older
 * community homicide is age-adjusted rate for 2013-2017 data combined, 39 (of 561) communities missing data, mostly do to cell suppression

 
 
```{r}
cbdLinkCA  <- read.csv(paste0(myPlace,"/myInfo/Tract to Community Linkage.csv"),colClasses = "character")  # file linking MSSAs to census 
comName    <- unique(cbdLinkCA[,c("comID","comName")])  


thisWork <-  left_join(tWork,comName,by="comName") %>%
              left_join(sdohComm,by=c("county","comID"))  %>%
                mutate(myGroup=ifelse(pValue < 0.05,ifelse(`Percent Change` > 0,"1 Increase","2 Decrease"),"3 Same"),
                       lessThanBachelors = 100- est_edu,
                       myLabel=paste(county,comName),
                       hRank = rank(-`Rate 2013-2017`),
                       top15 = ifelse(hRank < 16,"top15","notTot15"),
                       pov30plus = ifelse(est_pov > 30,"highPov","notHighPov")
                       ) 


top15stats <- thisWork %>% group_by(top15) %>% summarise(n=n(),
                                                         minPov=min(est_pov),
                                                         maxPov=max(est_pov),
                                                         meanPov=mean(est_pov)
                                                         )

pov30stats <- thisWork %>% filter(!is.na(`Rate 2013-2017`)) %>% mutate(
                               gState      = ifelse(`Rate 2013-2017` > 5.295,1,0),
                               gStateTwice = ifelse(`Rate 2013-2017` > 2*5.295,1,0) 
) %>% group_by(pov30plus) %>%   
  summarise(n=n(),
  aboveState = sum(gState),
  wayAboveState = sum(gStateTwice)
  )
  

qplot(thisWork$est_pov,
      geom="histogram",
      binwidth = 2,  
      main = "Histogram - Community Poverty", 
      xlab = "Percent of Community Below Federal Poverty Level",  
      fill=I("blue"), 
      col=I("red"),
      xlim=c(0,100))

qplot(thisWork$lessThanBachelors,
      geom="histogram",
      binwidth = 2,  
      main = "Histogram - Community Education", 
      xlab = "Percent of Community LESS than Bachelors degree",  
      fill=I("blue"), 
      col=I("red"),
      xlim=c(0,100))


qplot(thisWork$`Rate 2013-2017`,
      geom="histogram",
      binwidth = 0.5,  
      main = "Histogram - Community Homicide Rate", 
      xlab = "Homicde Rate",  
      fill=I("blue"), 
      col=I("red"))
```


#### observations:
 * poverty - normal-ish, slight right skew (can't be smaller than 0 or larger than 100) 
 * education - skewed left  
 * homicide -  strong skew right


<br><br>


### now looking at the "raw" associations 


```{r}
ggplot(thisWork, aes(x =est_pov, y = `Rate 2013-2017`) ) +   geom_point()  
```

 * appears to have strong linear association, but "spread" increasing with poverty (transform?)




```{r}
ggplot(thisWork, aes(x =lessThanBachelors, y = `Rate 2013-2017`) ) +   geom_point()  
```

 * non-linear relationship - log-linear (i.e. exponential) explore transformation; 
 * explore difference in pattern between poverty and education with homicide
 
<br>


####  check that this is not a coding error with eduction by comapring "our" ACS data pull to HPI

```{r}

ggplot(thisWork, aes(x = bachelorsed, y = est_edu) ) +
    geom_point()  
```

 * not a coding error (lack of exact correlation is from different years of data)




### "added value" exploration with:
 * linear modeling lines
 * color based on CHANGES in homicide rates
 * dot size based on number of homicides
 * very messy pop-up to get county/community name


```{r}
thisWork <- filter(thisWork,!is.na(`Ndeaths 2013-2017`))
```

<br><br><br>


```{r}
# mytext=paste("County = ", thisWork$county.x, "\n" , "Number of Homicides = ", thisWork$`Ndeaths 2013-2017`,"\n" , "Homicide Rate = ", thisWork$`Rate 2013-2017`,  "\n" ,"Poverty: ",thisWork$est_pov,"\n" ,"Row Number: ",rownames(thisWork),sep="")    


sRate <-  unlist(filter(t.dat,county=="CALIFORNIA",yearG3=="2015-2017",CAUSE=="E03")  %>%   select(aRate) )


 
 
 p <- ggplot(thisWork, aes(x = est_pov, y = `Rate 2013-2017`,size=`Ndeaths 2013-2017`,label=myLabel)) +
    guides(size=guide_legend(title="Number of Deaths")) +
     geom_point(aes(color = factor(myGroup)),alpha=.8)  +
    guides(color=guide_legend(title="Increase/Decrease")) +
     geom_smooth(method='lm') +
    geom_hline(yintercept =  sRate,color="orange",linetype = "dashed") +
     labs(title="Scatterplot of Community Povery Level by Coummunity Homicide Rate, 2013-2017") +
       ylab("age-adjusted Rate per 100,000") +
      xlab("Percent of Community Below Federal Poverty") +
         guides(fill=FALSE)
 


p



#pp=plotly_build(p)   
#style( pp, text=mytext, traces = c(1, 2, 3) )
```


<br><br><br>


```{r}
ggplotly(p)
```

<br><br><br>




```{r}
# mytext=paste("County = ", thisWork$county.x, "\n" , "Number of Homicides = ", thisWork$`Ndeaths 2013-2017`,"\n" , "Homicide Rate = ", thisWork$`Rate 2013-2017`,  "\n" ,"Poverty: ",thisWork$est_pov,"\n" ,"Row Number: ",rownames(thisWork),sep="")    









 
 p <- ggplot(thisWork, aes(x = lessThanBachelors, y = `Rate 2013-2017`,size=`Ndeaths 2013-2017`,label=myLabel)) +
    guides(size=guide_legend(title="Number of Deaths")) +
     geom_point(aes(color = factor(myGroup)),alpha=.8)  +
    guides(color=guide_legend(title="Increase/Decrease")) +
     geom_smooth(method='lm') +
    geom_hline(yintercept =  sRate,color="orange",linetype = "dashed") +
     labs(title="Scatterplot of % in Community < Bachelors Degreee by Homicide Rate, 2013-2017") +
       ylab("age-adjusted Rate per 100,000") +
      xlab("Percent of Community Below Bachelors") +
         guides(fill=FALSE)
 


p






#pp=plotly_build(p)   
#style( pp, text=mytext, traces = c(1, 2, 3) )
```


<br><br><br>

```{r}
ggplotly(p)
```



```{r}

## WRITE
write_csv(thisWork,"forJN_5_SDOH_and_Homicide_community.csv")

```



